Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-03-06 manuscript/manuscript.tex
updated manuscript

\documentclass[twocolumn, 11pt, draft]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=raggedright,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[numbers]{natbib}
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

@@

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. It does so by jointly considering the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six network methods to discover new biomarkers for breast cancer susceptibility by searching subnetworks with higher association scores to this phenotype. We interrogate a familial breast cancer genome-wide association study focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value = $7.8 \times 10^{-5}$ ) and occupy more central positions in the network than average. Additionally, it includes subnetworks of mechanisms related to cancer, like protein folding (\emph{HSPA1A}, \emph{HSPA1B}, and \emph{HSPA1L}) or mitochondrial ribosomes (\emph{MRPS30}, \emph{MRPS31}, \emph{MRPS18B}). We also observed a general dysregulation in the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. By trading statistical astringency for biological meaningfulness, most network methods get more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. 
\end{abstract}

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, the experimental setting also presents intrinsic challenges. Some of them stem from the high-dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with large and moderate effects can be detected. And it is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically Bonferroni. However, Bonferroni is known to be overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in a disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

@@

Network-based, biomarker discovery methods exploit the aforementioned strategy to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each SNP has a score of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Different solutions have been proposed to this problem, often stemming from divergent different mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \cite{gwinner_network-based_2016}, which focuses on star subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, focus on interconnected genes with high association with the disease. However, they differ in their tolerance to the inclusion of low scoring nodes, and the possible number of disconnected subnetworks in the solution. Lastly, other methods also consider the topology of the network, favoring solutions that are densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

In this work, we analyze the effectiveness of these six network methods for biomarker discovery on GWAS data. While all of them capture susceptibility mechanisms resembling that postulated by the omnigenic model, they do so in different ways, and provide a representative view of the field. We focus on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study of familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network-based methods to recover additional breast cancer biomarkers. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

@@

The GENE Sisters (GENESIS) study was designed to investigate risk factors for familial breast cancer in the French population \cite{sinilnikova_genesis:_2016}. Index cases are patients with infiltrating mammary or ductal adenocarcinoma, who had a sister with breast cancer, and who have been tested negative for \emph{BRCA1} and \emph{BRCA2} pathogenic variants. Controls are unaffected colleagues and/or friends of the cases, born around the year of birth of the corresponding case (\textpm{} 3 years). We focused on the 2\,577 samples of European ancestry, of which 1\,279 are controls and 1\,298 are cases. The genotyping was performed using the iCOGS array, a custom Illumina array designed to study genetic susceptibility of hormone-related cancers \cite{sakoda_turning_2013}. It contains 211\,155 SNPs, including SNPs putatively associated with breast, ovarian, and prostate cancers, SNPs associated with survival after diagnosis, and SNPs associated to other cancer-related traits, as well as functional candidate variants in selected genes and pathways.

@@

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy - Weinberg equilibrium in controls (P-value \textless 0.001), and those missing on more than 10\% of the samples. A subset of 20 duplicated SNPs in \emph{FGFR2} were also removed. In addition, we removed the samples with more than 10\% missing genotypes. After control for relatedness, 17 additional samples were removed (6 for sample identity error, 6 false ``friend/control'' having family link with other samples, 3 ``friend/control'' having a high relatedness score). Lastly, based on study selection criteria, 11 other samples were removed (1 control having cancer, 4 index cases with no affected sister, 3 half-sisters, 1 sister with CLIS, 1 with \emph{BRCA1/2} mutation detected in the family, 1 with molecular diagnosis not received). The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

We looked for population structure that could create confounding associations. A PCA revealed no visual differential population structure between cases and controls (Supplementary Figure \ref{sfig:pcs}). Independently, we did not find evidence of genomic inflation (\(\lambda\) = 1.05) either, further confirming the absence of confounding population structure.

\subsection{High-score subnetwork search algorithms}
\subsubsection{SNP and gene association}
\label{methods:node_score}
To measure association between a genotype and the phenotype, we performed a per-SNP 1~d.f.~\(\chi\)\textsuperscript{2} allelic test using PLINK v1.90 \cite{chang_second-generation_2015}. Then, we used VEGAS2v2 to compute the gene-level association score \cite{mishra_vegas2:_2015} from the SNP P-values. In order to map SNPs to genes we used their overlap on the genome: all SNPs located within the boundaries of a gene, \textpm 50 kb, were mapped to that gene. To compute the gene association score we used the 10\% of SNPs linked to the gene with lowest P-values. We used the 62\,193 genes described in GENCODE 31 \cite{frankish_gencode_2019}, although only 54\,612 could be mapped to at least one SNP. Out of those, we focused exclusively on the 32\,767 that had a gene symbol. Out of the SNPs 197\,083 remaining after quality control, 164\,037 were mapped to at least one of these genes.

We use such mapping to compare the outputs of methods that produce SNP- to those that produce gene-lists, and vice versa. In the former case, we consider any gene that can be mapped to any of the selected SNPs as selected as well. In the latter, we consider all the SNPs that can be mapped to that gene as selected by the method.

\subsubsection{Mathematical notation}
\label{methods:notation}
In this article, we use undirected, vertex-weighted networks, or graphs, $G = (V,E,w)$. $V = \{v_{1}, \dots{}, v_{n}\}$ refers to the vertices, with weights $w: V \rightarrow \mathbb{R}$. Equivalently, $E \subseteq \{\{x,y\} | x,y \in V \wedge x \neq y\}$ refers to the edges. When referring to a subnetwork S, $V_{S}$ is the set of nodes in S and $E_{S}$ is the set of edges in S. A special case of subgraphs are \emph{connected} subgraphs, which occur when every node in the subgraph can be reached from any other node.

On top of a weight, nodes have other properties provided by the topology of the graph. In this article we focus on two: degree centrality, and betweenness centrality. The degree centrality, or degree, is the number of edges that a node has. The betweenness centrality, or betweenness, is the number of times a node participates in the shortest paths between two other nodes.

@@

\subsubsection{Methods used}
\label{methods:methods}

\begin{table}[htbp]
  \caption{\label{tab:method_comparison} Summary of the differences between the studied algorithms.}
  \centering
  \begin{threeparttable}
    \begin{tabular}{l|llllllll}
      Method & Field & Nodes & Exhaustive & Solution & Components & Input & Scoring & Reference\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Module & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Module & \(\ge\) 1 & Genotypes & \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Module & 1 & Summary & -log\textsubscript{10}(P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      \item \textbf{\emph{Field}}: field in which the algorithm was developed.\\
      \item \textbf{\emph{Nodes}}: the type of network, either gene (protein-protein interaction network usually) or a SNP network.\\
      \item \textbf{\emph{Exhaustive}}: whether all the possible solutions given the selected hyperparameters are explored.\\
      \item \textbf{\emph{Solution}}: additional properties are enforced on the solution subnetwork, other than being dense in high scores and connected.\\
      \item \textbf{\emph{Components}}: number of connected subnetworks in the solution.\\
      \item \textbf{\emph{Input}}: genotype data or GWAS summary statistics.\\
      \item \textbf{\emph{Scoring}}: how SNP/gene P-values are transformed into node scores.\\
      \item \textbf{\emph{Reference}}: original publication featuring the algorithm.
    \end{tablenotes}
  \end{threeparttable}
\end{table}

Beyond the assumption that genes that contribute to the same function will be nearby in the protein-protein interaction network (PPIN), they might be topologically related to each other in diverse ways (densely interconnected modules, nodes around a hub, a path, etc.). That is not the only choice to make: how to score the nodes, whether the affected mechanisms form a single connected component or several, how to frame the problem in a computationally efficient fashion, what is the best network to use, etc. In consequence, multiple solutions have been proposed. In this article, we examine six of them: five that explore the PPIN, and one which explores SNP networks. We selected methods that were open source, had an implementation available, and an accessible documentation. Their main differences are summarized in Table \ref{tab:method_comparison}.

\begin{description}
\item[{dmGWAS}] dmGWAS searches the subgraph with the highest local density in low P-values \cite{jia_dmgwas:_2011}. To that end it searches candidate subnetwork solutions using a greedy, ``seed and extend'', heuristic:

@@

\begin{equation*} 
Z_m = \frac{\sum_{i \in S} z_i}{\sqrt{k}}
\end{equation*}

where \emph{k} is the number of genes in the subgraph; z\textsubscript{i} is the score of gene $i$, computed as \(\phi\)\textsuperscript{-1}(1 - P-value\textsubscript{i}); and \(\phi\)\textsuperscript{-1} is the inverse normal distribution function.
\item Identify neighboring nodes i.e. nodes at distance \(\le\) \emph{d}. We set d = 2.
\item Add the neighboring nodes whose inclusion increases the Z\textsubscript{m+1} more than Z\textsubscript{m} \texttimes{} (1 + r). In our experiments, we set r = 0.1.
\item Repeat 2-4 until no increment Z\textsubscript{m} \texttimes{} (1 + r) is possible.
\end{enumerate}

@@

where Z\textsubscript{m}(\(\pi\)) represent a vector containing 100\,000 random subsets of the same number of genes.

@@

were p(v) = w(v) - w' is the \emph{profit} of adding a node, c(e) = w' is the \emph{cost} of adding an edge, and $w' = min_{v \in V_{G}} w(v)$. All three are positive quantities. heinz implements the algorithm from \citet{ljubic_algorithmic_2006}, which in practice is often fast and optimal, neither is guaranteed. We used BioNet's implementation of heinz, available on Bioconductor \cite{beisser_bionet:_2010,heinz}.

\item[{HotNet2}] HotNet2 was developed to find connected subgraphs of genes frequently mutated in cancer \cite{leiserson_pan-cancer_2015}. To that end, it considers both the local topology of the network and the scores of the nodes. The former is captured by an insulated heat diffusion process: at the beginning, the score of the node determines its initial heat; iteratively each node yields heat to its ``colder'' neighbors, and receives heat from its ``hotter'' neighbors, while retaining part of its own (hence, \emph{insulated}). This process continues until equilibrium is reached, and results in a similarity matrix F. F is used to compute the similarity matrix E that accounts also for similarities in node scores as

@@

where $\operatorname{diag}(w(V))$ is a diagonal matrix with the node scores in its diagonal. We scored the nodes as in \citet{nakka_gene_2016}, assigning a score of 0 for the genes with low probability of being associated to the disease, and -log\textsubscript{10}(P-value) to those likely to be. In this dataset, the threshold separating both was a P-value of 0.125, which was obtained using a local FDR approach \cite{scheid_twilight;_2005}. To obtain densely connected subnetworks, HotNet2 prunes E, only preserving edges such that w(E) \textgreater{} \(\delta\). Lastly, HotNet2 evaluates the statistical significance of the subnetworks by comparing their size to the size of networks obtained by permuting the node scores. HotNet2 has two parameters: the restart probability \(\beta\), and the threshold heat \(\delta\). Both parameters are set automatically by the algorithm, and are robust \cite{leiserson_pan-cancer_2015}. HotNet2 is implemented in Python \cite{hotnet2}.

@@

where \(\lambda\) and \(\eta\) are hyperparameters that control the sparsity and the connectivity of the model. Given two hyperparameters, the aforementioned problem has a unique solution, that SConES finds using a graph min-cut procedure. We used the version on SConES implemented in the R package martini \cite{martini}. As in \citet{azencott_efficient_2013}, we selected \(\lambda\) and \(\eta\) by cross-validation, choosing the values that produce the most stable solution across folds. Note that the solution to the above problem can consist of several connected subnetworks which are disconnected from each other. In this case, the selected hyperparameters were \(\eta\) = 3.51, \(\lambda\) = 210.29 for SConES GS; \(\eta\) = 3.51, \(\lambda\) = 97.61 for SConES GM; and \(\eta\) = 3.51, \(\lambda\) = 45.31 for SConES GI.
\end{description}

\begin{description}
\item[{SigMod}] SigMod aims at identifying the highest-scoring, most densely connected gene subnetwork \cite{liu_sigmod:_2017}. It addresses an optimization problem similar to that of SConES (Equation \ref{eq:scones}), but using the adjacency matrix rather than the Laplacian matrix (Section \ref{methods:notation}), to quantify solutions containing many edges.

@@

In practice, and despite their similarities and their involvement in cancer mechanisms, the solutions are remarkably different (Supplementary Figure \ref{sfig:pearson_methods}A). That is due to the particularities of the methods which directly or indirectly provide information about the dataset. For instance, the fact that LEAN did not provide any biomarkers implies that there is no gene such that both itself and its environment are on average strongly associated with the disease.

>>>>

\documentclass[twocolumn, 11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=justified,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[numbers]{natbib}
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

@@

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. It does so by jointly considering the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six network methods to discover new biomarkers for breast cancer susceptibility by searching subnetworks with higher association scores to this phenotype. We interrogate a familial breast cancer genome-wide association study focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical astringency for biological meaningfulness, most network methods get more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value~=~$7.8 \times 10^{-5}$ ) and occupy more central positions in the network than average. Additionally, it includes subnetworks of mechanisms related to cancer, like protein folding (\emph{HSPA1A}, \emph{HSPA1B}, and \emph{HSPA1L}) or mitochondrial ribosomes (\emph{MRPS30}, \emph{MRPS31}, \emph{MRPS18B}). We also observed a general dysregulation in the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. 
\end{abstract}

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high-dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with large and moderate effects can be detected. And it is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically Bonferroni. However, Bonferroni is known to be overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

@@

Network-based, biomarker discovery methods exploit this relatedness to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each SNP has a score of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Different solutions have been proposed to this problem, often stemming from divergent mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \cite{gwinner_network-based_2016}, which focuses on \emph{star} subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, focus on larger subnetworks interconnected genes with high association scores. However, they differ in their tolerance to the inclusion of low scoring nodes, and the possible number of disconnected subnetworks in the solution. Lastly, other methods also consider the topology of the network, favoring solutions that are not only high scoring, but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

In this work, we analyze the effectiveness of these six network methods for biomarker discovery on GWAS data. While all of them capture susceptibility mechanisms resembling that postulated by the omnigenic model, they do so in different ways, and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network-based methods to recover additional breast cancer biomarkers. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

@@

The GENE Sisters (GENESIS) study was designed to investigate risk factors for familial breast cancer in the French population \cite{sinilnikova_genesis:_2016}. Index cases are patients with infiltrating mammary or ductal adenocarcinoma, who had a sister with breast cancer, and who have been tested negative for \emph{BRCA1} and \emph{BRCA2} pathogenic variants. Controls are unaffected colleagues and/or friends of the cases, born around the year of birth of their corresponding case (\textpm{} 3 years). We focused on the 2\,577 samples of European ancestry, of which 1\,279 are controls and 1\,298 are cases. The genotyping was performed using the iCOGS array, a custom Illumina array designed to study genetic susceptibility of hormone-related cancers \cite{sakoda_turning_2013}. It contains 211\,155 SNPs, including SNPs putatively associated with breast, ovarian, and prostate cancers, SNPs associated with survival after diagnosis, and SNPs associated to other cancer-related traits, as well as functional candidate variants in selected genes and pathways.

@@

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy - Weinberg equilibrium in controls (P-value \textless~0.001), and those missing on more than 10\% of the samples. A subset of 20 duplicated SNPs in \emph{FGFR2} were also removed. In addition, we removed the samples with more than 10\% missing genotypes. After control for relatedness, 17 additional samples were removed (6 for sample identity error, 6 controls related to other samples, and 3 additional controls having a high relatedness score). Lastly, based on study selection criteria, 11 other samples were removed (1 control having cancer, 4 index cases with no affected sister, 3 half-sisters, 1 sister with CLIS, 1 with a \emph{BRCA1/2} mutation detected in the family, 1 with unknown molecular diagnosis). The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

We looked for population structure that could produce spurious associations. A PCA revealed no visual differential population structure between cases and controls (Supplementary Figure~\ref{sfig:pcs}). Independently, we did not find evidence of genomic inflation (\(\lambda\)~=~1.05) either, further confirming the absence of confounding population structure.

\subsection{High-score subnetwork search algorithms}
\subsubsection{SNP and gene association}
\label{methods:node_score}
To measure association between a genotype and the phenotype, we performed a per-SNP 1~d.f.~\(\chi\)\textsuperscript{2} allelic test using PLINK v1.90 \cite{chang_second-generation_2015}. Then, we used VEGAS2v2 \cite{mishra_vegas2:_2015} to compute the gene-level association score from the P-values of the SNPs mapped to them. Specifically, for each gene we only used the 10\% of SNPs mapped to it with lowest P-values. We mapped SNPs to genes through their genomic coordinates: all SNPs located within the boundaries of a gene, \textpm 50 kb, were mapped to that gene. We used the 62\,193 genes described in GENCODE 31 \cite{frankish_gencode_2019}, although only 54\,612 could be mapped to at least one SNP. Out of those, we focused exclusively on the 32\,767 that had a gene symbol. Out of the 197\,083 SNPs remaining after quality control, 164\,037 were mapped to at least one of these genes.

We use such mapping to compare the outputs of methods that produce SNP-lists to those that produce gene-lists, and vice versa. For the former, we consider any gene that can be mapped to any of the selected SNPs as selected as well. For the latter, we consider all the SNPs that can be mapped to that gene as selected by the method.

\subsubsection{Mathematical notation}
\label{methods:notation}
In this article, we use undirected, vertex-weighted networks, or graphs, $G = (V,E,w)$. $V = \{v_{1}, \dots{}, v_{n}\}$ refers to the vertices, with weights $w: V \rightarrow \mathbb{R}$. Equivalently, $E \subseteq \{\{x,y\} | x,y \in V \wedge x \neq y\}$ refers to the edges. When referring to a subnetwork $S$, $V_{S}$ is the set of nodes in $S$ and $E_{S}$ is the set of edges in $S$. A special case of subgraphs are \emph{connected} subgraphs, which occur when every node in the subgraph can be reached from any other node.

On top of a weight, nodes have other properties, provided by the topology of the graph. In this article we focus on two: degree centrality, and betweenness centrality. The degree centrality, or degree, is the number of edges that a node has. The betweenness centrality, or betweenness, is the number of times a node participates in the shortest paths between two other nodes.

@@

\subsubsection{Networks used}
\label{methods:networks}

\paragraph{Gene-gene network}
Out of the six methods tested, five use a gene-gene interaction network (Section~\ref{methods:methods}). Although their respective statistical frameworks are compatible with any type of network (protein interactions, gene coexpression, regulatory, etc.), for practical reasons we focused on a protein-protein interaction network (PPIN), as they are interpretable, well characterized, and most of the methods were designed to scale appropriately to it. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless specified otherwise, we used only interactions coming from high-throughput experiments, leaving out targeted studies that might bias the topology of the network. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we were able to map 142\,541 to a pair of gene symbols. The scoring function for the nodes changed from method to method (Section~\ref{methods:methods}).

Additionally, we compared the results of the aforementioned PPIN with those obtained on another PPIN built using interactions coming from both high-throughput and targeted studies. In that case, out of the 179\,332 interactions in HINT, 173\,797 mapped to a pair of gene symbols.

\paragraph{SNP networks}
SConES \cite{azencott_efficient_2013} is the only of the studied methods designed to handle SNP networks. As in gene networks, two SNPs are linked in a SNP network when there is evidence of shared functionality between two SNPs. \citet{azencott_efficient_2013} proposed three ways of building these networks: connecting the SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists, on top of GM (``GI network''). We focused on the GI network, as it fits the scope of this article better, using the PPIN described above. However, at different stages of this work we also used GS and GM for comparison. For the GM network, we used the mapping described in Section~\ref{methods:node_score}. In all three the node scores are the association score of the individual SNPs with the phenotype (1~d.f.~\(\chi\)\textsuperscript{2}).
\subsubsection{Methods used}
\label{methods:methods}

\begin{table}[htbp]
  \caption{\label{tab:method_comparison} Summary of the differences between the studied algorithms.}
  \centering
  \begin{threeparttable}
    \begin{tabular}{l|llllllll}
      Method & Field & Nodes & Exhaustive & Solution & Comp. & Input & Scoring & Ref.\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Module & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Module & \(\ge\) 1 & Genotypes & 1 d.f. \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Module & 1 & Summary & -log\textsubscript{10}(P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      \footnotesize{
        \item \textbf{\emph{Field}}: field in which the algorithm was developed. \textbf{\emph{Nodes}}: the type of network, either gene (protein-protein interaction network usually) or a SNP network. \textbf{\emph{Exhaustive}}: whether all the possible solutions given the selected hyperparameters are explored. \textbf{\emph{Solution}}: additional properties are enforced on the solution subnetwork, other than being dense in high scores and connected. \textbf{\emph{Comp.}}: number of connected subnetworks in the solution. \textbf{\emph{Input}}: genotype data or GWAS summary statistics. \textbf{\emph{Scoring}}: how SNP/gene P-values are transformed into node scores. \textbf{\emph{Ref.}}: original publication featuring the algorithm.
      }
    \end{tablenotes}
  \end{threeparttable}
\end{table}

Genes that contribute to the same function are nearby in the PPIN, might be topologically related to each other in diverse ways (densely interconnected modules, nodes around a hub, a path, etc.). How to account for that is not the only choice a developer of network methods needs to make: how to score the nodes, whether the affected mechanisms form a single connected component or several, how to frame the problem in a computationally efficient fashion, which network to use, etc. In consequence, multiple solutions have been proposed. In this article, we examine six of them: five that explore the PPIN, and one which explores SNP networks. We selected methods that were open source, had an implementation available, and an accessible documentation. Their main differences are summarized in Table~\ref{tab:method_comparison}.

\begin{description}
\item[{dmGWAS}] dmGWAS seeks the subgraph with the highest local density in low P-values \cite{jia_dmgwas:_2011}. To that end it searches candidate subnetwork solutions using a greedy, ``seed and extend'', heuristic:

@@

\begin{equation*} 
Z_m = \frac{1}{\sqrt{k}} \sum_{i \in S} z_i
\end{equation*}

where \emph{k} is the number of genes in the subgraph; z\textsubscript{i} is the Z score of gene $i$, computed as \(\phi\)\textsuperscript{-1}(1 - P-value\textsubscript{i}); and \(\phi\)\textsuperscript{-1} is the inverse normal distribution function.
\item Identify neighboring nodes of $S$ i.e. nodes at distance \(\le\) d. We set d~=~2.
\item Add the neighboring nodes whose inclusion increases the Z\textsubscript{m+1} more than a threshold Z\textsubscript{m}~\texttimes{}~(1 + r). In our experiments, we set r~=~0.1.
\item Repeat 2-4 until no further enlargement is possible.
\end{enumerate}

@@

where Z\textsubscript{m}(\(\pi\)) represents a vector containing 100\,000 random subsets of the same number of genes.

@@

were p(v)~=~w(v) - w' is the \emph{profit} of adding a node, c(e)~=~w' is the \emph{cost} of adding an edge, and $w' = min_{v \in V_{G}} w(v)$. All three are positive quantities. Heinz implements the algorithm from \citet{ljubic_algorithmic_2006}, which although in practice is often fast and optimal, neither is guaranteed. We used BioNet's implementation of heinz, available on Bioconductor \cite{beisser_bionet:_2010,heinz}.

\item[{HotNet2}] HotNet2 was developed to find connected subgraphs of genes frequently mutated in cancer \cite{leiserson_pan-cancer_2015}. To that end, it considers both the local topology of the network and the scores of the nodes. The former is captured by an insulated heat diffusion process: at the beginning, the score of the node determines its initial heat; iteratively each node yields heat to its ``colder'' neighbors, and receives heat from its ``hotter'' neighbors, while retaining part of its own (hence, \emph{insulated}). This process continues until equilibrium is reached, and results in a similarity matrix $F$. $F$ is used to compute the similarity matrix $E$ that accounts also for similarities in node scores as

@@

where $\operatorname{diag}(w(V))$ is a diagonal matrix with the node scores in its diagonal. We scored the nodes as in \citet{nakka_gene_2016}, assigning a score of 0 for the genes with low probability of being associated to the disease, and -log\textsubscript{10}(P-value) to those likely to be. In this dataset, the threshold separating both was a P-value of 0.125, which was obtained using a local FDR approach \cite{scheid_twilight;_2005}. To obtain densely connected subnetworks, HotNet2 prunes $E$, only preserving edges such that $w(E) > \delta$. Lastly, HotNet2 evaluates the statistical significance of the subnetworks by comparing their size to the size of networks obtained by permuting the node scores. HotNet2 has two parameters: the restart probability \(\beta\), and the threshold heat \(\delta\). Both parameters are set automatically by the algorithm, and are robust \cite{leiserson_pan-cancer_2015}. HotNet2 is implemented in Python \cite{hotnet2}.

@@

where \(\lambda\) and \(\eta\) are hyperparameters that control the sparsity and the connectivity of the model. Given two hyperparameters, the aforementioned problem has a unique solution, that SConES finds using a graph min-cut procedure. We used the version on SConES implemented in the R package martini \cite{martini}. As in \citet{azencott_efficient_2013}, we selected \(\lambda\) and \(\eta\) by cross-validation, choosing the values that produce the most stable solution across folds. Note that the solution to the above problem can consist of several connected subnetworks which are disconnected from each other. In this case, the selected hyperparameters were \(\eta\)~=~3.51, \(\lambda\)~=~210.29 for SConES GS; \(\eta\)~=~3.51, \(\lambda\)~=~97.61 for SConES GM; and \(\eta\)~=~3.51, \(\lambda\)~=~45.31 for SConES GI.
\end{description}

\begin{description}
\item[{SigMod}] SigMod aims at identifying the highest-scoring, most densely connected gene subnetwork \cite{liu_sigmod:_2017}. It addresses an optimization problem similar to that of SConES (Equation \ref{eq:scones}), but using the adjacency matrix rather than the Laplacian matrix (Section~\ref{methods:notation}), to quantify solutions containing many edges.

@@

In practice, and despite their similarities and their involvement in cancer mechanisms, the solutions are remarkably different (Supplementary Figure~\ref{sfig:pearson_methods}A). That is due to the particularities of the methods which directly or indirectly provide information about the dataset. For instance, the fact that LEAN did not provide any biomarkers implies that there is no gene such that both itself and its environment are on average strongly associated with the disease.
